``Interestingness'' in Mathematical Theory Exploration

### 2018-04-06 PhDThesis/outline.tex
Thesis Nix changes

Title: Use of "Interestingness" in Scaling Mathematical Theory Exploration

Introduction:

 - Description of problem
 - Motivation

Background (/Literature review)

 - History of automated mathematical discovery
 - History of what counts as "interesting"
 - Relevance filtering

Contributions:

 - Empirical proposal for measuring what is "interesting" (TEBenchmark)
 - Evaluation and comparison of existing systems using this benchmark (QuickSpec
   v1 and IsaCoSy)

Application of machine learning to focus existing systems on what is "interesting"

 - Overview of machine learning techniques:
  - Supervised vs unsupervised
  - Clustering
  - Representing structured inputs
 - Bucketing algorithm
  - Use with pseudorandom similarity metric
  - Use with "flattened tree" similarity metric
  - Potentially some other comparison?

Practical uses of automated mathematical discovery systems

 - Optimisation
  - Self-optimising programs
  - Use with Hutter search?
 - Finding abstractions (anti-unification)
  - Concept formation?

Application of discovery system to existing software packages

 - Packages from Hackage (may need extra Arbitrary instances)
 - GHC libraries (base, maybe more)
 - Maybe use test suites as ground truth?

Conclusions

---

We have a few things sat around gathering dust. Can we fit them in somewhere?

 - Purely-functional self-modifying code. This could be an implementation
   vehicle for open-ended self-optimising code.

 - Powerplay. This uses self-invented problems. Could we use "interesting"
   conjectures from a theory exploration system to guide the self-improvement of
   an automated theorem prover?

Things which have influenced my thinking, which would be nice to include:

 - Interestingness based on compression progress. The "compression" could
   perhaps come from writing proof tactics? For example, try to prove any given
   conjecture using a single 'prove' tactic. Given a (conjecture, tactic) pair,
   and an existing tactic, we deem the conjecture to be "interesting" if it
   can't be proved by the existing tactic, it can be proved by the new tactic,
   and the new tactic is smaller than the old tactic. Hmm... This wouldn't allow
   tactics to increase in size. The compression progress idea relies on a
   lossless compression of all 'past experience', which in this situation would
   just be a list of all conjectures we've tried. This just passes the buck on
   what's "interesting", since those conjectures need to be chosen somehow...

 - Artificial curiosity systems, like problem generators rewarded by the
   variance of how difficult a population of solvers found the problem to be.

 - Recursive self-improvement: take an algorithm which searches for improvements
   to itself, and run it for a few self-improvement iterations. How do the
   improvements to the search algorithm compare against the difficulty of
   improving an optimised system ('recalcitrance')?

 - Can we create a compiler by providing a property discovery algorithm with a
   specification and an model interpreter (e.g. for x86 machine code)?

 - A result I'd really like to have is a system which collapses down abstraction
   layers. This would be especially nice if it worked across languages, and we
   could either use staged execution/compilation and/or a database of
   compiled/transformed code.

>>>>

\title{``Interestingness'' in Mathematical Theory Exploration}
\documentclass{article}
\begin{document}
\maketitle

\begin{itemize}
\item Introduction
  \begin{itemize}
  \item Description of problem (simple examples)
  \item Motivation (education, research, formalising, verification, ...)
  \item Difficulties (ambiguity, open-endedness, large search space, ...)
  \end{itemize}

\item Contributions:
  \begin{itemize}
  \item TEBenchmark (large, unambiguous empirical measurement and comparison)
  \item MLSpec (bucketing methods, recurrent clustering, etc.)
  \end{itemize}

\item Background/Literature review
  \begin{itemize}
  \item History of automated discovery/creativity
    \begin{itemize}
    \item Theorem proving (Logic Theorist)
    \item Theory formation/exploration (AM, HR, Graffiti, IsaCoSy, IsaScheme,
      QuickSpec, Speculate, ...)
    \item Scientific discovery (Adam/Eve, invariant discovery)
    \item Artificial curiosity (model building, motor babbling, play, language invention, ...)
    \item Artistic creativity?
    \end{itemize}
  \item History of "interestingness"
    \begin{itemize}
    \item From philosophical perspective (surprising, novel, useful, Wundt
      curves, ...)
    \item From real implementations (non-redundancy, precision/recall,
      compression progress, learnability, ...)
    \end{itemize}
  \item Overview of machine learning techniques:
    \begin{itemize}
    \item Models (neural networks, generative models, probabilistic programming,
      gradient descent, kernel methods, population models, ...)
    \item Supervised (classification, reinforcement learning, ...)
    \item Unsupervised (clustering, co-evolution, adversarial, ...)
    \item Feature engineering
      \begin{itemize}
      \item Hand-crafted features (comparison with probabalistic programming?)
      \item Distributed representations
      \item Learned representations (auto-encoding, deep learning, convolutional
        neural networks, ...)
      \end{itemize}
    \item Structured representations (recurrent NNs, recursive NNs, tree
      kernels, Tree Based CNN, generative models, ...)
    \end{itemize}
  \item (Statistical) machine learning in formal systems
    \begin{itemize}
    \item Relevance filtering (Sledgehammer, Josef Urban's alternatives)
    \item Recurrent clustering
    \item Program manipulation (Learning to Execute, Tree Based CNN tasks
      (program classification, bubble-sort identification), auto-generated bug
      fixes, auto-generated compiler test cases, ...
    \end{itemize}
  \end{itemize}

\item TEBenchmark (mostly from paper)
  \begin{itemize}
    \item Methodology
    \item TIP corpus
    \item Haskell translation (specific issues, etc.)
    \item Isabelle translation (specific issues, etc.)
    \item QuickSpec analysis
    \item IsaCoSy analysis
    \item QuickSpec/IsaCoSy comparison
    \item Discussion of pros and cons
  \end{itemize}

\item Bucketing (mostly paper and transfer report)
  \begin{itemize}
  \item Justification (exponential running times, large numbers of discarded
    equations, low precision, ...)
  \item Bucketing as a solution (big-O analysis, oracles, ...)
  \item Bucketing algorithms
    \begin{itemize}
    \item Pseudorandom features (baseline)
    \item Circular convolution (language agnostic, distributed representation,
      not learned)
    \item Recurrent clustering (Haskell-specific, not distributed, hand-crafted,
      solves global name problem)
    \item Non-recurrent clustering?
    \end{itemize}
  \item Application to TEBenchmark (impact on total recall)
  \item Discussion
    \begin{itemize}
    \item Pros and cons
    \end{itemize}
  \end{itemize}

\item Other contributions (may splice into the other sections)
  \begin{itemize}
    \item Sandboxed evaluation of Haskell code (nix-eval)
    \item Haskell AST extraction and annotation (GHC plugin, standalone command)
    \item Exploration of popular Hackage packages
  \end{itemize}

\item Conclusions
  \begin{itemize}
  \item Exploration is an important problem with many applications
  \item Hampered by ambiguous goals, small sample sizes, etc.
  \item Hampered by (lack of) efficiency
  \item Unambiguous benchmark is provided for comparison
  \item ``Smart'' bucketing method to improve efficiency
  \item Lots of room for future work (other corpora, other tools, other
    languages, applications of results, optimisation through profiling,
    abstraction invention through anti-unification, self-modelling, ...)
  \end{itemize}
\end{itemize}

\end{document}
